---
title: "Data 624 Project 1"
author: "Steven Gonzalez"
date: "3/30/2025"
output:
  html_document: default
  pdf_document: default
---
```{r message=FALSE, warning=FALSE}
library(fpp3)
library(ggplot2)
library(knitr)
library(openxlsx)
library(readxl)
library(tidyverse)
library(VIM)
```
## Prompt A
Part A – ATM Forecast, ATM624Data.xlsx

In part A, I want you to forecast how much cash is taken out of 4 different ATM machines for May 2010. The data is given in a single file. The variable ‘Cash’ is provided in hundreds of dollars, other than that it is straight forward. I am being somewhat ambiguous on purpose to make this have a little more business feeling. Explain and demonstrate your process, techniques used and not used, and your actual forecast. I am giving you data via an excel file, please provide your written report on your findings, visuals, discussion and your R code via an RPubs link along with the actual.rmd file. Also please submit the forecast which you will put in an Excel readable file.

## Approach


## Data Exploration and Preparation


### Load and View Data
```{r message=FALSE, warning=FALSE}
training <- read.csv("https://raw.githubusercontent.com/Stevee-G/Data624/refs/heads/main/Project2/TrainingData.csv")
testing <- read.csv("https://raw.githubusercontent.com/Stevee-G/Data624/refs/heads/main/Project2/TestData.csv")

str(training)
str(testing)
```

As can be seen above, the data sets are comprised of 33 variables each, 32 of them most likely being predictor variables. The training data set has 2571 observations while the testing data set has 267 observations.

### Address Formatting and Missing Values
```{r message=FALSE, warning=FALSE}

```

### Prepare Tsibbles
```{r message=FALSE, warning=FALSE}

```

## ATM 1
```{r message=FALSE, warning=FALSE}

```

## ATM 2
```{r message=FALSE, warning=FALSE}

```

## ATM 3
```{r message=FALSE, warning=FALSE}

```

## ATM 4
```{r message=FALSE, warning=FALSE}

```


## Conclusion

