---
title: "Data 624 Homework 1"
author: "Steven Gonzalez"
date: "2/9/2025"
output:
  html_document: default
  pdf_document: default
---

## Load Packages
```{r message=FALSE}
library(fpp3)
library(USgas)
```

## Exercise 1
Explore the following four time series: `Bricks` from `aus_production`, `Lynx` from `pelt`, `Close` from `gafa_stock`, `Demand` from `vic_elec`.

Use `?` (or `help()`) to find out about the data in each series.
What is the time interval of each series?
Use `autoplot()` to produce a time plot of each series.
For the last plot, modify the axis labels and title.
```{r message=FALSE}
?aus_production
aus_production #used to get further familiarized with the data
```
As can be seen from the results above, the `Bricks` time series from `aus_production` has a quarterly time interval. Below is the time plot illustrating this using `autoplot()`.
```{r message=FALSE, warning=FALSE}
autoplot(aus_production, Bricks)
```

```{r message=FALSE}
?pelt
pelt #used to get further familiarized with the data
```
As can be seen from the results above, the `Lynx` time series from `pelt` has an annual time interval. Below is the time plot illustrating this using `autoplot()`.
```{r message=FALSE, warning=FALSE}
autoplot(pelt, Lynx)
```

```{r message=FALSE}
?gafa_stock
gafa_stock #used to get further familiarized with the data
```
As can be seen from the results above, the `Close` time series from `gafa_stock` has a time interval with specific dates that seem to be business days, which would make sense given that it is a data set on stock prices. Below is the time plot illustrating this using `autoplot()`.
```{r message=FALSE, warning=FALSE}
autoplot(gafa_stock, Close)
```

```{r message=FALSE}
?vic_elec
vic_elec #used to get further familiarized with the data
```
As can be seen from the results above, the `Demand` time series from `vic_elec` has a half-hourly time interval. Below is the time plot illustrating this using `autoplot()` with modified title and axis labels.
```{r message=FALSE, warning=FALSE}
autoplot(vic_elec, Demand) +
  labs(title = "Half-hourly electricity demand for Victoria, Australia",
       y = "Total Electricity Demand (MWh)")
```

## Exercise 2
Use `filter()` to find what days corresponded to the peak closing price for each of the four stocks in `gafa_stock`.
```{r}
aapl_peak <- gafa_stock %>% 
  filter(Symbol == "AAPL") %>% 
  select(Symbol, Date, Close) %>%
  slice_max(Close, n = 1)
aapl_peak

amzn_peak <- gafa_stock %>% 
  filter(Symbol == "AMZN") %>% 
  select(Symbol, Date, Close) %>%
  slice_max(Close, n = 1)
amzn_peak

fb_peak <- gafa_stock %>% 
  filter(Symbol == "FB") %>% 
  select(Symbol, Date, Close) %>%
  slice_max(Close, n = 1)
fb_peak

goog_peak <- gafa_stock %>% 
  filter(Symbol == "GOOG") %>% 
  select(Symbol, Date, Close) %>%
  slice_max(Close, n = 1)
goog_peak
```

## Exercise 3
Download the file `tute1.csv` from the book website, open it in Excel (or some other spreadsheet application), and review its contents. You should find four columns of information. Columns B through D each contain a quarterly series, labelled Sales, AdBudget and GDP. Sales contains the quarterly sales for a small company over the period 1981-2005. AdBudget is the advertising budget and GDP is the gross domestic product. All series have been adjusted for inflation.

You can read the data into R with the following script:
`tute1 <- readr::read_csv("tute1.csv")
View(tute1)`

Convert the data to time series
`mytimeseries <- tute1 |>
  mutate(Quarter = yearquarter(Quarter)) |>
  as_tsibble(index = Quarter)`

Construct time series plots of each of the three series
`mytimeseries |>
  pivot_longer(-Quarter) |>
  ggplot(aes(x = Quarter, y = value, colour = name)) +
  geom_line() +
  facet_grid(name ~ ., scales = "free_y")`
Check what happens when you don’t include `facet_grid()`.
```{r}
url <- "https://raw.githubusercontent.com/Stevee-G/Data624/refs/heads/main/tute1.csv"
tute1 <- readr::read_csv(url) #Had to modify the command in order to make the RMD reproducible
View(tute1)

mytimeseries <- tute1 %>% 
  mutate(Quarter = yearquarter(Quarter)) %>% 
  as_tsibble(index = Quarter) #Modified the pipe due to personal preference

mytimeseries %>% 
  pivot_longer(-Quarter) %>% 
  ggplot(aes(x = Quarter, y = value, colour = name)) +
  geom_line() +
  facet_grid(name ~ ., scales = "free_y")

mytimeseries %>% 
  pivot_longer(-Quarter) %>% 
  ggplot(aes(x = Quarter, y = value, colour = name)) +
  geom_line()
```

## Exercise 4
The `USgas` package contains data on the demand for natural gas in the US.

Install the `USgas` package.
Create a `tsibble` from `us_total` with year as the index and state as the key.
Plot the annual natural gas consumption by state for the New England area (comprising the states of Maine, Vermont, New Hampshire, Massachusetts, Connecticut and Rhode Island).
```{r}
#USgas package was installed and loaded in a previous section
?us_total
glimpse(us_total)

us_total_ts <- us_total %>%
  as_tsibble(index = year, key = state)

new_england <- us_total_ts %>% 
  filter(state == "Maine" |
           state == "Vermont" |
           state == "New Hampshire" |
           state == "Massachusetts" |
           state == "Connecticut" |
           state == "Rhode Island")

autoplot(new_england, y)
```

## Exercise 5
Download `tourism.xlsx` from the book website and read it into R using `readxl::read_excel()`.
Create a tsibble which is identical to the `tourism` tsibble from the `tsibble` package.
Find what combination of `Region` and `Purpose` had the maximum number of overnight trips on average.
Create a new tsibble which combines the Purposes and Regions, and just has total trips by State.
```{r message=FALSE, warning=FALSE}
url1 <- "https://raw.githubusercontent.com/Stevee-G/Data624/refs/heads/main/tourism.csv"
tourism1 <- readr::read_csv(url1) #Had to resort to csv due to an issue with OneDrive making the excel unreadable when importing data from online
glimpse(tourism1)

tourism #Take a look at the tourism tsibble in order to compare with tsibble made from the tourism excel sheet
?tourism #Get familiar with tourism tsibble to identify index

tourism1_ts <- tourism1 %>%
  mutate(Quarter = yearquarter(Quarter)) %>% 
  as_tsibble(index = Quarter, key = c(Region, State, Purpose))
tourism1_ts #Glimpse and compare tourism1_ts to tourism tsibble
```
By comparing the tsibbles produced above we can say for certain that the new `tourism1_ts` is identical to the original `tourism`.
```{r}
max_avg_trips <- tourism1_ts %>% 
  group_by(Region, Purpose) %>% 
  summarise(avg_trips = mean(Trips)) %>% 
  slice_max(avg_trips, n = 1) %>% 
  arrange(desc(avg_trips))
max_avg_trips
```
Through the code chunk above, we can see that the combination of `Region` and `Purpose` with the maximum number of overnight trips on average was "Melbourne" and "Visiting".
```{r}
total_trips <- tourism1_ts %>%
  group_by(State) %>% #By using the group_by function on State, we can collapse all region and purpose data to see what the total trips are by state/territory
  summarise(tot_trips = sum(Trips))
total_trips
```

## Exercise 8
Use the following graphics functions: `autoplot()`, `gg_season()`, `gg_subseries()`, `gg_lag()`, `ACF()` and explore features from the following time series: “Total Private” `Employed` from `us_employment`, `Bricks` from `aus_production`, `Hare` from `pelt`, “H02” `Cost` from `PBS`, and `Barrels` from `us_gasoline`.

Can you spot any seasonality, cyclicity and trend?
What do you learn about the series?
What can you say about the seasonal patterns?
Can you identify any unusual years?
```{r warning=FALSE}
us_employment
private_employment <- us_employment %>% 
  filter(Title == "Total Private")
autoplot(private_employment, Employed)
gg_season(private_employment, y = Employed)
gg_subseries(private_employment, y = Employed)
gg_lag(private_employment, y = Employed)
ACF(private_employment, y = Employed) %>% 
  autoplot()
```
As can be seen above, the “Total Private” `Employed` time series from `us_employment` does seem to show some seasonality where the numbers employed go up through the middle months of the year just to drop again towards the end of the year but still higher than they were in the beginning of the year. The series did have some cyclicity throughout the 80 year stretch that could be due to fluctuations in the economy, especially the unusual dip recorded in the few years leading up to 2010. The over all trend of the series is positive.
```{r warning=FALSE}
aus_production
autoplot(aus_production, Bricks)
gg_season(aus_production, y = Bricks)
gg_subseries(aus_production, y = Bricks)
gg_lag(aus_production, y = Bricks)
ACF(aus_production, y = Bricks) %>% 
  autoplot()
```
As can be seen above, the `Bricks` time series from `aus_production` gives clear signs of seasonality where the number of bricks produced goes up in quarters two and three to drop dramatically towards the beginning of the following year. The series was cyclical every few years and had some especially hard dips around the years 1975, 1983, 1991, and 1996. The over all trend began positive up until around 1983 where things began going south. This could probably be due to consumer demand for bricks decreasing throughout the last few decades.
```{r warning=FALSE}
pelt
autoplot(pelt, Hare)
#gg_season(pelt, y = Hare, period = "year") Couldn't get it to work :(
gg_subseries(pelt, y = Hare)
gg_lag(pelt, y = Hare)
ACF(pelt, y = Hare) %>% 
  autoplot()
```
As can be seen above, the `Hare` time series from `pelt` shows very little, f any, signs of seasonality. Seasonality can only be defined within a year, whether it be quarterly, monthly, or weekly. This time series does not show to have yearly patterns but rather a pattern that seems to last around 10 years at a time. This pattern can be better described as cyclical. With regards to trend, there seems to be no indication of one, whether positive or negative. This series is best defined as a pattern of habitual decade long cycles repeating nonstop.
```{r warning=FALSE}
PBS
H02 <- PBS %>%
  filter(ATC2 == "H02")
autoplot(H02, Cost)
gg_season(H02, y = Cost)
gg_subseries(H02, y = Cost)
#gg_lag(H02, y = Cost) Couldn't get it to work :(
ACF(H02, y = Cost) %>% 
  autoplot()
```
As can be seen above, the “H02” `Cost` time series from `PBS` is seasonal for some `Concession`/`Type` combinations but not for all. Right off the bat we can see in the `gg_season` and `gg_subseries` plots that general/co_payments combination is not seasonal, while the rest are. Concessional/co-payments seems to rise throughout the middle months, concessional/safety net seems to dip during the middle months, and general/safety net seems to rise slightly during the final months of the year. Given the above plots, we can't say for sure whether there is a cycle since the fluctuations seen can be explained by the seasonality mentioned earlier. There are noticeable trends with concessional/co_payments and concessional/safety net moving in the positive throughout the years while general_co-payments and general/safety net remain about the same.
```{r warning=FALSE}
us_gasoline
autoplot(us_gasoline, Barrels)
gg_season(us_gasoline, y = Barrels)
gg_subseries(us_gasoline, y = Barrels)
gg_lag(us_gasoline, y = Barrels)
ACF(us_gasoline, y = Barrels) %>% 
  autoplot()
```
As can be seen above, the `Barrels` time series from `us_gasoline` does seem to show slight seasonality with the barrels sold between the months of March and October bumping up and going back down shortly after. There don't seem to be any prominent cycles except for the dip observed around the years 2007 to 2013 which was most likely caused by the [economic recession of 2008](https://www.history.com/topics/21st-century/recession). The overall trend, however, is positive and it seems that US gasoline consumption managed to recover from the downturn.

## Works Cited
History.com Editors. (2019, October 11). Great Recession. HISTORY; A&E Television Networks. https://www.history.com/topics/21st-century/recession

