---
title: "Data 624 Homework 1"
author: "Steven Gonzalez"
date: "2/9/2025"
output: pdf_document
---

## Load Packages
```{r message=FALSE}
library(fpp3)
```

## Exercise 1
Explore the following four time series: `Bricks` from `aus_production`, `Lynx` from `pelt`, `Close` from `gafa_stock`, `Demand` from `vic_elec`.

Use `?` (or `help()`) to find out about the data in each series.
What is the time interval of each series?
Use `autoplot()` to produce a time plot of each series.
For the last plot, modify the axis labels and title.
```{r message=FALSE}
?aus_production
aus_production #used to get further familiarized with the data
```
As can be seen from the results above, the `Bricks` time series from `aus_production` has a quarterly time interval. Below is the time plot illustrating this using `autoplot()`.
```{r message=FALSE, warning=FALSE}
autoplot(aus_production, Bricks)
```

```{r message=FALSE}
?pelt
pelt #used to get further familiarized with the data
```
As can be seen from the results above, the `Lynx` time series from `pelt` has an annual time interval. Below is the time plot illustrating this using `autoplot()`.
```{r message=FALSE, warning=FALSE}
autoplot(pelt, Lynx)
```

```{r message=FALSE}
?gafa_stock
gafa_stock #used to get further familiarized with the data
```
As can be seen from the results above, the `Close` time series from `gafa_stock` has a time interval with specific dates that seem to be business days, which would make sense given that it is a data set on stock prices. Below is the time plot illustrating this using `autoplot()`.
```{r message=FALSE, warning=FALSE}
autoplot(gafa_stock, Close)
```

```{r message=FALSE}
?vic_elec
vic_elec #used to get further familiarized with the data
```
As can be seen from the results above, the `Demand` time series from `vic_elec` has a half-hourly time interval. Below is the time plot illustrating this using `autoplot()` with modified title and axis labels.
```{r message=FALSE, warning=FALSE}
autoplot(vic_elec, Demand) +
  labs(title = "Half-hourly electricity demand for Victoria, Australia",
       y = "Total Electricity Demand (MWh)")
```

## Exercise 2
Use `filter()` to find what days corresponded to the peak closing price for each of the four stocks in `gafa_stock`.
```{r}
aapl_peak <- gafa_stock %>% 
  filter(Symbol == "AAPL") %>% 
  select(Symbol, Date, Close) %>%
  slice_max(Close, n = 1)
aapl_peak

amzn_peak <- gafa_stock %>% 
  filter(Symbol == "AMZN") %>% 
  select(Symbol, Date, Close) %>%
  slice_max(Close, n = 1)
amzn_peak

fb_peak <- gafa_stock %>% 
  filter(Symbol == "FB") %>% 
  select(Symbol, Date, Close) %>%
  slice_max(Close, n = 1)
fb_peak

goog_peak <- gafa_stock %>% 
  filter(Symbol == "GOOG") %>% 
  select(Symbol, Date, Close) %>%
  slice_max(Close, n = 1)
goog_peak
```

## Exercise 3